### 2022-12-09 abstract.tex
Submitted for alan review

    In the later chapters we focus on probabilistic programming and developing theory and tools
    to automate statistical inference using randomized algorithms based on
    Monte Carlo Markoc chain (MCMC) and variational inference (VI).
    \Cref{ch:lic} considers parameterizing $q(x)$ by graph neural networks which condition each
    node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
    \cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
    the length of execution traces.
    When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
    considers the problem of tail anisotropy through both a theoretical and practical perspective.
    However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
    the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
    probabilistic program's source code.

>>>>

    In the later chapters we focus on probabilistic programming and developing theory and tools
    to automate statistical inference using randomized algorithms based on
    Monte Carlo Markov chain (MCMC) and variational inference (VI).
    \Cref{ch:lic} considers parameterize $q(x)$ by graph neural networks which condition each
    node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
    \cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
    the length of execution traces.
    When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
    considers the problem of tail anisotropy through both a theoretical and practical perspective.
    However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
    the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
    probabilistic program's source code.

### 2022-12-09 design/continuous_nips/continuous.tex
Submitted for alan review

Classical statistical learning theory asserts that to achieve generalization
one must use training sample size that sufficiently exceeds the complexity of
the learning model, where the latter is typically represented by the number of
parameters \citep[or some related structural parameter; see][]{HFT09}.  In particular,
this seems to suggest the conventional wisdom that one should not use models
that fit the training data exactly.  However, modern machine learning practice
often seems to go against this intuition, using models with so many parameters
that the training data can be perfectly interpolated, in which case the
training error vanishes. It has been shown that models such as deep neural
networks, as well as certain so-called interpolating kernels and decision
trees, can generalize well in this regime. In particular,
\cite{BHMM19} empirically demonstrated a phase transition in generalization
performance of learning models which occurs at an \emph{interpolation
  thershold}, i.e., a point where training error goes to zero (as one varies the
ratio between the model complexity and the sample size). Moving away from this
threshold in either direction tends to reduce the generalization error, leading
to the so-called \emph{double descent} curve.

@@

We next give a few simple examples to provide some intuition. First, note
that every $1\times 1$ random matrix is determinant preserving simply
because taking a determinant is an identity transfomation in one
dimension. Similarly, every fixed matrix is determinant preserving because
in this case taking the expectation is an identity
transformation. In all other cases, however, Definition \ref{d:main}
has to be verified more carefully. Further examples (positive and
negative) follow.
\begin{example}
  If $\A$ has i.i.d. Gaussian entries $a_{ij}\sim\Nc(0,1)$, then
  $\A$ is d.p.~because $\E[\det(\A)]=0$.
\end{example}

@@

The proof comes in three steps: (i) for fixed $z > 0$, consider the limiting behavior of $ \delta(z) \equiv \tr (\X \X^\top/n + z \I_n)^{-1}/n$ as $n,d \to \infty$ and state
\begin{equation}\label{eq:limit2}
  \lim_{n,d \to \infty} \delta(z) - m(z) \to 0
\end{equation}
almost surely for some $m(z)$ to be defined; (ii) show that both $\delta(z)$ and its derivate $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take $z \to 0^+$ in \eqref{eq:limit2} and state
\begin{equation}\label{eq:limit3}
  \lim_{z \to 0^+} \lim_{n,d \to \infty} \delta(z) - \lim_{z \to 0^+} m(z) \to 0
\end{equation}
almost surely, given that the limit $\lim_{z \to 0^+} m(z) \equiv m(0)$ exists and eventually (iii) exchange the two limits in \eqref{eq:limit3} with Moore-Osgood theorem, to reach
\[
  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left(\frac1n \X \X^\top + z \I_n \right)^{-1} - m(0) \to 0.
\]

>>>>

Classical statistical learning theory asserts that to achieve generalization
one must use training sample size that sufficiently exceeds the complexity of
the learning model, where the latter is typically represented by the number of
parameters \citep[or some related structural parameter; see][]{HFT09}.  In particular,
this seems to suggest the conventional wisdom that one should not use models
that fit the training data exactly.  However, modern machine learning practice
often seems to go against this intuition, using models with so many parameters
that the training data can be perfectly interpolated, in which case the
training error vanishes. It has been shown that models such as deep neural
networks, as well as certain so-called interpolating kernels and decision
trees, can generalize well in this regime. In particular,
\cite{BHMM19} empirically demonstrated a phase transition in generalization
performance of learning models which occurs at an \emph{interpolation
  threshold}, i.e., a point where training error goes to zero (as one varies the
ratio between the model complexity and the sample size). Moving away from this
threshold in either direction tends to reduce the generalization error, leading
to the so-called \emph{double descent} curve.

@@

We next give a few simple examples to provide some intuition. First, note
that every $1\times 1$ random matrix is determinant preserving simply
because taking a determinant is an identity transformation in one
dimension. Similarly, every fixed matrix is determinant preserving because
in this case taking the expectation is an identity
transformation. In all other cases, however, Definition \ref{d:main}
has to be verified more carefully. Further examples (positive and
negative) follow.
\begin{example}
  If $\A$ has i.i.d. Gaussian entries $a_{ij}\sim\Nc(0,1)$, then
  $\A$ is d.p.~because $\E[\det(\A)]=0$.
\end{example}

@@

The proof comes in three steps: (i) for fixed $z > 0$, consider the limiting behavior of $ \delta(z) \equiv \tr (\X \X^\top/n + z \I_n)^{-1}/n$ as $n,d \to \infty$ and state
\begin{equation}\label{eq:limit2}
  \lim_{n,d \to \infty} \delta(z) - m(z) \to 0
\end{equation}
almost surely for some $m(z)$ to be defined; (ii) show that both $\delta(z)$ and its derivative $\delta'(z)$ are uniformly bounded (by some quantity independent of $z>0$) so that by Arzela-Ascoli theorem, $\delta(z)$ converges uniformly to its limit and we are allowed to take $z \to 0^+$ in \eqref{eq:limit2} and state
\begin{equation}\label{eq:limit3}
  \lim_{z \to 0^+} \lim_{n,d \to \infty} \delta(z) - \lim_{z \to 0^+} m(z) \to 0
\end{equation}
almost surely, given that the limit $\lim_{z \to 0^+} m(z) \equiv m(0)$ exists and eventually (iii) exchange the two limits in \eqref{eq:limit3} with Moore-Osgood theorem, to reach
\[
  \lim_{n,d \to \infty} \lim_{z \to 0^+} \frac1n \tr \left(\frac1n \X \X^\top + z \I_n \right)^{-1} - m(0) \to 0.
\]

### 2022-12-09 design/equivalents_nips/neurips_2020.tex
Submitted for alan review

The effectiveness of sketching has also been extensively studied in
the context of second order optimization. These methods differ
depending on how the sketch is applied to the Hessian matrix, and
whether or not it is applied to the gradient as well. The class of
methods discussed in Section \ref{s:newton}, including Randomized
Subspace Newton and the Generalized Kaczmarz method, relies on
projecting the Hessian downto a low-dimensional subspace, which makes our
results directly applicable. A related family of methods uses
the so-called Iterative Hessian Sketch (IHS) approach \citep{pilanci2016iterative,lacotte2019faster}. The
similarities between IHS and the Subspace Newton-type methods (see
\citep{Qu2015Feb} for a comparison) suggest that our techniques could be
extended to provide precise convergence guarantees also to the IHS.
Finally, yet another family of Hessian sketching methods has been studied by
\citep{roosta2019sub,sketched-ridge-regression,XRM17_theory_TR,YXRM18_TR,fred_newtonMR_TR,distributed-newton,determinantal-averaging}.
These methods  preserve the  rank of the
Hessian, and so their convergence
guarantees do not rely on the residual projection.

@@

We complement the results of Section \ref{s:experiments} with
empirical results on four additional libsvm datasets \citep{libsvm} (bringing
the total number of benchmark datasets to eight), which further
establish the accuracy of our surrogate expressions for the low-rank approximation
error. Similary as in Figure \ref{f:nystrom}, we use the sketched
Nystr\"om method \citep{revisiting-nystrom} with the RBF kernel
$k(\a_i,\a_j)=\exp(-\|\a_i-\a_j\|^2/(2\sigma^2))$, for several values
of the parameter $\sigma$. The values of $\sigma$ were chosen so as to
demonstrate the effectiveness of our theoretical predictions both when the
stable rank is moderately large and when it is very small.

>>>>

The effectiveness of sketching has also been extensively studied in
the context of second order optimization. These methods differ
depending on how the sketch is applied to the Hessian matrix, and
whether or not it is applied to the gradient as well. The class of
methods discussed in Section \ref{s:newton}, including Randomized
Subspace Newton and the Generalized Kaczmarz method, relies on
projecting the Hessian down to a low-dimensional subspace, which makes our
results directly applicable. A related family of methods uses
the so-called Iterative Hessian Sketch (IHS) approach \citep{pilanci2016iterative,lacotte2019faster}. The
similarities between IHS and the Subspace Newton-type methods (see
\citep{Qu2015Feb} for a comparison) suggest that our techniques could be
extended to provide precise convergence guarantees also to the IHS.
Finally, yet another family of Hessian sketching methods has been studied by
\citep{roosta2019sub,sketched-ridge-regression,XRM17_theory_TR,YXRM18_TR,fred_newtonMR_TR,distributed-newton,determinantal-averaging}.
These methods  preserve the  rank of the
Hessian, and so their convergence
guarantees do not rely on the residual projection.

@@

We complement the results of Section \ref{s:experiments} with
empirical results on four additional libsvm datasets \citep{libsvm} (bringing
the total number of benchmark datasets to eight), which further
establish the accuracy of our surrogate expressions for the low-rank approximation
error. Similarly as in Figure \ref{f:nystrom}, we use the sketched
Nystr\"om method \citep{revisiting-nystrom} with the RBF kernel
$k(\a_i,\a_j)=\exp(-\|\a_i-\a_j\|^2/(2\sigma^2))$, for several values
of the parameter $\sigma$. The values of $\sigma$ were chosen so as to
demonstrate the effectiveness of our theoretical predictions both when the
stable rank is moderately large and when it is very small.

### 2022-12-09 ftvi/ftvi_icml/ftvi_icml.tex
Submitted for alan review

Note that $\overline{\cL^p_\alpha} \supset \cE^q_\beta$ for all $p,q,\alpha,\beta$, so \Cref{thm:distn_class_closed}
by itself does not preclude transformations of fat-tailed base distributions to light-tailed targets.
Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushfoward,
hence heavy-to-light transformations are impossible. Note here there is no ascending union over
tail parameters (i.e., $\cL^p_\alpha$ instead of $\overline{\cL^p_\alpha}$).

@@

\begin{proof}[Proof of \Cref{corr:closure_polynomials}]
  Let $X \in \cE^p_\alpha$.
  By considering sufficiently large $X$ such that leading powers dominate, it suffices to consider monomials $Y = X^k$.
  Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointness of $\mathcal{E}$ and $\mathcal{L}$.
\end{proof}

@@

For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
is comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empkkirically observed
this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
the absence of outliers when using a Gaussian base distribution resulting in more stable ELBO gradients. This suggests
other techniques for handling outliers such as winsorization may also be helpful, and we leave further investigation
for future work.

@@

To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021 \citep{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicid (CMS) claims during 2008--2010 \citep{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events. Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}. Both datasets exhibited superior fits after allowing for heavier tails, with a further improved fit using ATAF for the CMS claims dataset.
%Using identical flow architectures and optimizers, the log likelihoods in \Cref{tab:density-estimation} suggest that models permitting tail-anisotropy (ATAF) yield significantly superior fits.

>>>>

Note that $\overline{\cL^p_\alpha} \supset \cE^q_\beta$ for all $p,q,\alpha,\beta$, so \Cref{thm:distn_class_closed}
by itself does not preclude transformations of fat-tailed base distributions to light-tailed targets.
Under additional assumptions on $f_\theta$, we further establish a partial converse that a fat-tailed base distribution's tail parameter is unaffected after pushforward,
hence heavy-to-light transformations are impossible. Note here there is no ascending union over
tail parameters (i.e., $\cL^p_\alpha$ instead of $\overline{\cL^p_\alpha}$).

@@

\begin{proof}[Proof of \Cref{corr:closure_polynomials}]
  Let $X \in \cE^p_\alpha$.
  By considering sufficiently large $X$ such that leading powers dominate, it suffices to consider monomials $Y = X^k$.
  Notice $\PP(Y \geq x) = \PP(X \geq x^{1/k}) = \Theta(e^{-\alpha x^{p/k}})$, and so
  $Y \in \cE^{p/k}_\alpha$. The result follows by disjointedness of $\mathcal{E}$ and $\mathcal{L}$.
\end{proof}

@@

For \cref{tab:diamonds} and \cref{tab:eight_schools}, the flow transform $\Phi_{\text{Flow}}$ used for ADVI, TAF, and ATAF
is comprised of two hidden layers of 32 units each. NUTS uses no such flow transform. Variational parameters for each normalizing flow were initialized
using \texttt{torch}'s default Kaiming initialization \citep{he2015delving} Additionally, the tail parameters $\nu_i$
used in ATAF were initialized to all be equal to the tail parameters learned from training TAF. We empirically observed
this resulted in more stable results (less variation in ELBO / $\log p(y)$ across trials), which may be due to
the absence of outliers when using a Gaussian base distribution resulting in more stable ELBO gradients. This suggests
other techniques for handling outliers such as winsorization may also be helpful, and we leave further investigation
for future work.

@@

To examine the advantage of tail-anisotropic modelling in practice, we considered two benchmark datasets from financial (daily log returns for five industry indices during 1926--2021 \citep{fama2015five}) and actuarial (per-patient inpatient and outpatient cumulative Medicare/Medicaid (CMS) claims during 2008--2010 \citep{cms}) applications where practitioners actively seek to model fat-tails and account for black-swan events. Identical flow architectures and optimizers were used in both cases, with log-likelihoods presented in \Cref{tab:density-estimation}. Both datasets exhibited superior fits after allowing for heavier tails, with a further improved fit using ATAF for the CMS claims dataset.
%Using identical flow architectures and optimizers, the log likelihoods in \Cref{tab:density-estimation} suggest that models permitting tail-anisotropy (ATAF) yield significantly superior fits.

### 2022-12-09 hta/neurips_2022.tex
Submitted for alan review

Continuing this line of work, here we
we consider how to statically analyze a probabilistic program in order to automate the inference of
tail behavior for any
random variables present. At present, correct inference of tail behaviour for target distributions remains an outstanding issue \citep{yao2018yes,wang2018variational}, which causes challenges for downstream Monte Carlo tasks. For example, importance sampling estimators can exhibit infinite variance if the tail of the approximating density is lighter than the target. Most prominant black-box variational inference methods are incapable of changing their tail behaviour from an initial proposal distribution \citep{jaini2020tails,ftvi}. MCMC algorithms may also lose ergodicity when the tail of the target density falls outside of a particular family \citep{roberts1996exponential}. All of these issues could be avoided if the tail of the target is known before runtime.

@@

For studying heavy tails, methods based on subexponential distributions
\citep{goldie1998subexponential} and generalized Pareto distributions (GPD) or
equivalently regularly varying distributions \citep{tajvidi2003confidence}
have recieved attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
Heavy tails can impact probabilistic machine learning methods in a number of ways.
The observation that density ratios $\frac{p(x)}{q(x)}$ tend to be heavy tailed
has resulted in new methods for smoothing importance sampling \citep{vehtari2015pareto},
adaptively modifying divergences \citep{wang2018variational}, and
diagnosing variational inference through the Pareto $\hat{k}$ diagnostic \citep{yao2018yes}.
These works are complementary to our paper and our reported results include $\hat{k}$
diagnostics for VI and $\hat\alpha$ tail index estimates based on GPD.

@@

There is a long history of formal methods and probabilistic programming
\cite{kozen1979semantics,jones1989probabilistic}. While much of the research
\cite{claret2013bayesian}
is concerned with defining formal semantics
and establishing invariants \cite{wang2018pmaf}
See \cite{bernstein2019static} for a recent review.
Static analysis utilizes the abstract syntax tree (AST)
representation of a program in order to compute invariants
(e.g. the return type of a function, the number of classes implementing a trait)
without executing the underlying program.
As dynamic analysis in PPs is less reliable due to non-determinism, static analysis
methods for PPs become increasingly important.

@@

In this section we demonstrate that GGA-based density estimation yields improvements across a variety of metrics.
We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushfowards of Normal distributions. To understand the individual
effect of using a GGA base distribution over standard normals versus more expressive pushforward maps \citep{durkan2019neural}, we also report ablation results where
normalizing flows are replaced by affine transforms as originally
proposed in \citep{kucukelbir2017automatic}. All experiments are repeated for 50 trials, trained to convergence using the Adam optimizer with manually tuned learning rate,
and conducted on i7-8700K CPU and GTX 1080 GPU hardware.

>>>>

Continuing this line of work, here we
we consider how to statically analyze a probabilistic program in order to automate the inference of
tail behavior for any
random variables present. At present, correct inference of tail behaviour for target distributions remains an outstanding issue \citep{yao2018yes,wang2018variational}, which causes challenges for downstream Monte Carlo tasks. For example, importance sampling estimators can exhibit infinite variance if the tail of the approximating density is lighter than the target. Most prominent black-box variational inference methods are incapable of changing their tail behaviour from an initial proposal distribution \citep{jaini2020tails,ftvi}. MCMC algorithms may also lose ergodicity when the tail of the target density falls outside of a particular family \citep{roberts1996exponential}. All of these issues could be avoided if the tail of the target is known before runtime.

@@

For studying heavy tails, methods based on subexponential distributions
\citep{goldie1998subexponential} and generalized Pareto distributions (GPD) or
equivalently regularly varying distributions \citep{tajvidi2003confidence}
have received attention historically. \citet{mikosch} presents closure theorems for regularly varying which are special cases of \Cref{prop:gga_add} and \Cref{lem:lipschitz}.
Heavy tails can impact probabilistic machine learning methods in a number of ways.
The observation that density ratios $\frac{p(x)}{q(x)}$ tend to be heavy tailed
has resulted in new methods for smoothing importance sampling \citep{vehtari2015pareto},
adaptively modifying divergences \citep{wang2018variational}, and
diagnosing variational inference through the Pareto $\hat{k}$ diagnostic \citep{yao2018yes}.
These works are complementary to our paper and our reported results include $\hat{k}$
diagnostics for VI and $\hat\alpha$ tail index estimates based on GPD.

@@

There is a long history of formal methods and probabilistic programming
\cite{kozen1979semantics,jones1989probabilistic}. While much of the research
\cite{claret2013bayesian}
is concerned with defining formal semantics
and establishing invariants \cite{wang2018pmaf}
See \cite{bernstein2019static} for a recent review.
Static analysis utilizes the abstract syntax tree (AST)
representation of a program in order to compute invariants
(e.g. the return type of a function, the number of classes implementing a trait)
without executing the underlying program.
As dynamic analysis in PPLs is less reliable due to non-determinism, static analysis
methods for PPLs become increasingly important.

@@

In this section we demonstrate that GGA-based density estimation yields improvements across a variety of metrics.
We consider the parametric family defined in \Cref{ssec:repr_dist} and compare against pushforwards of Normal distributions. To understand the individual
effect of using a GGA base distribution over standard normals versus more expressive pushforward maps \citep{durkan2019neural}, we also report ablation results where
normalizing flows are replaced by affine transforms as originally
proposed in \citep{kucukelbir2017automatic}. All experiments are repeated for 50 trials, trained to convergence using the Adam optimizer with manually tuned learning rate,
and conducted on i7-8700K CPU and GTX 1080 GPU hardware.

### 2022-12-09 ic/aistats2021/paper.tex
Submitted for alan review

\chapter{Accelerating Metropolis-hastings with lightweight inference compilation}
\label{ch:lic}

@@

A reference implementation for LIC and code to reproduce our experiments
have been made publically available\footnote{\url{https://github.com/facebookresearch/lightweight-inference-compilation}}.

@@

\Cref{fig:blr} shows the results of performing inference using LIC compared
against other MCMC inference methods. Results for existing IC approaches
\citep{le2017inference} are omitted because they are not comparable due to
lack of support for vector-valued random variables in the publically
available reference implementation \citep{pyprob2020}. All methods yield
similar predictive log-likelihoods on held-out test data, but LIC and NUTS
yield significantly higher ESS and $\widehat{R}$s closer to $1.0$ suggesting
better mixing and more effective sampling.

>>>>

\chapter{Accelerating Metropolis-Hastings with lightweight inference compilation}
\label{ch:lic}

@@

A reference implementation for LIC and code to reproduce our experiments
have been made publicly available\footnote{\url{https://github.com/facebookresearch/lightweight-inference-compilation}}.

@@

\Cref{fig:blr} shows the results of performing inference using LIC compared
against other MCMC inference methods. Results for existing IC approaches
\citep{le2017inference} are omitted because they are not comparable due to
lack of support for vector-valued random variables in the publicly
available reference implementation \citep{pyprob2020}. All methods yield
similar predictive log-likelihoods on held-out test data, but LIC and NUTS
yield significantly higher ESS and $\widehat{R}$s closer to $1.0$ suggesting
better mixing and more effective sampling.

### 2022-12-09 intro.tex
Submitted for alan review

When faced with difficult problems, randomized approximations are a tool
commonly employed by quantitative scientists which can offer algernative
algorithms and analyses. In this dissertation, we consider a range of problems
at the intersection of randomized methods and statistics.
Throughout our work, randomized methods are a unifying theme used in the first section to
construct tractable approximations and later as tools for automating computational Bayesian statistics.

@@

In the later chapters we focus on probabilistic programming and developing theory and tools
to automate statistical inference using randomized algorithms based on
Monte Carlo Markoc chain (MCMC) and variational inference (VI).
The second three chapters concern the setting of probabilistic programming, a computational tool
for specifying probabilistic models and automating inference through MCMC and VI. In particular, we
study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
importance sampling and variational inference.
In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
\cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
the length of execution traces.
When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
considers the problem of tail anisotropy through both a theoretical and practical perspective.
We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
an anisotropic approximation (fat-tailed variational inference, FTVI) where
an anisotropic product base measures is pushed forwards through a bijective neural network,
and confirm that in practice FTVI improves both density estimation as well as variational inference.
However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
such as sums and products. This enables a priori computation of tail parameters, which we show
improves the stability and convergence of a number of inference tasks.
These chapters correspond to the following publications:
\begin{itemize}
    \item \fullcite{lic}
    \item \fullcite{ftvi}
    \item \fullcite{gga}
\end{itemize}

>>>>

When faced with difficult problems, randomized approximations are a tool
commonly employed by quantitative scientists which can offer alternative
algorithms and analyses. In this dissertation, we consider a range of problems
at the intersection of randomized methods and statistics.
Throughout our work, randomized methods are a unifying theme used in the first section to
construct tractable approximations and later as tools for automating computational Bayesian statistics.

@@

In the later chapters we focus on probabilistic programming and developing theory and tools
to automate statistical inference using randomized algorithms based on
Monte Carlo Markov chain (MCMC) and variational inference (VI).
The second three chapters concern the setting of probabilistic programming, a computational tool
for specifying probabilistic models and automating inference through MCMC and VI. In particular, we
study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
importance sampling and variational inference.
In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
\cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
the length of execution traces.
When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
considers the problem of tail anisotropy through both a theoretical and practical perspective.
We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
an anisotropic approximation (fat-tailed variational inference, FTVI) where
an anisotropic product base measures is pushed forwards through a bijective neural network,
and confirm that in practice FTVI improves both density estimation as well as variational inference.
However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
such as sums and products. This enables a priori computation of tail parameters, which we show
improves the stability and convergence of a number of inference tasks.
These chapters correspond to the following publications:
\begin{itemize}
    \item \fullcite{lic}
    \item \fullcite{ftvi}
    \item \fullcite{gga}
\end{itemize}
